---
title: "Task1"
author: "Sofia"
date: "10/01/2024"
format:
  html:
    embed-resources: true
editor: source
---

# 1.1Atomic Types
Question description

Create the following variables related to your chosen object:
- 4 numeric values (e.g., population, temperature, size, rating)  
- 3 integer values (e.g., number of moons, buildings, inhabitants)  
- 2 logical values (e.g., has_water, is_inhabited)  
- 3 character values (e.g., name, region, category)

```{r}
# code for solving task
#integer 
population <- 39800
area <- 98  # square kilometers
districts <- 4

#charakter 
city_name <- "Zvolen"
city_old_name <- "Zolna"
region <- "Banska Bystrica Region" 

#numeric
average_temperature <- 10.1       # °C
city_budget_million <- 142.3      # million euros
annual_rainfall_mm <- 720.5       # mm per year
green_area_percent <- 36.8        # % of city covered by greenery

#logical 
has_hospital <- TRUE
is_capital <- FALSE
```



# 1.2Vectors
Create three different vectors:

1. **Numeric vector** combining all your numeric and integer variables.  
   *(Use `c()`.)*

2. **Sequence vector** from any number between 10–50 to 100, Choose step between 2 and 10.  
   *(Use `seq()` or `:`.)*

3. **Logical vector** of length 10 with random `TRUE`/`FALSE` values using `sample()`.

```{r}
#_code for solving task
# 1. Numeric vector
numeric_vector <- c(population, area, districts, average_temperature,
                    city_budget_million, annual_rainfall_mm, green_area_percent)
numeric_vector


# 2. Sequence vector
sequence_vector <- seq(5, 100, by = 20)
sequence_vector




# 3. Logical vector
logical_vec <- sample(c(TRUE, FALSE), size = 10, replace = TRUE)
logical_vec

```
### 1.3 MATRIX
Create a **3×3 matrix** with random numbers between **-50 and 50**  
(using `sample()` or `runif()`), and name its **rows and columns** according to your theme.





```{r}
#_code for solving task
set.seed(2024)
my_matrix <- matrix(sample(-50:50, 9, replace = TRUE), nrow = 3, ncol = 3)
rownames(my_matrix) <- c("Center", "East", "West")
colnames(my_matrix) <- c("Economy", "Education", "Tourism")
my_matrix

rowMeans(my_matrix)
colMeans(my_matrix)

```
# 1.4 LIST
Create a **list** that combines all your previous objects:  
atomic variables, vectors, and matrix.
Add one extra element — a short **description** of your universe as a string.
```{r}
#_code for solving task
zvolen_list <- list(
  city_name = city_name,
  region = region,
  population = population,
  numeric_vector = numeric_vector,
  sequence_vector = sequence_vector,
  logical_vec = logical_vec,
  matrix = my_matrix,
  description = "This is the data universe of Zvolen a historic Slovak city full of green spaces, education, and culture."
)
zvolen_list

```



```{r}

```

















# 1.5 Factor
Create a **factor variable** representing categories within your universe, e.g.:












- `"low"`, `"medium"`, `"high"`  
- `"infant"`, `"child"`, `"adult"`  
- `"poor"`, `"average"`, `"rich"`

Make sure you have **at least 5 observations** total.  
Then use `table()` to show how many objects fall into each category.
```{r}
#code_for solving task
development_stage <- factor(c("low", "medium", "high", "medium", "high", "low", "medium"))
table(development_stage)
```
# 1.6 Data Frame
Create a **data frame** related to your object that includes:

1. **5–8 observations** (rows)  
2. **At least 3 columns** of different data types  
   (e.g., name, numeric value, logical flag, category)
   dka
   dsds
   
   
   
   
   
   
   
   
   
   
```{r}
#_code for solving task
zvolen_df <- data.frame(
  District = c("Center", "East", "West", "North", "South"),
  Population = c(12000, 9500, 7800, 5500, 7000),
  Has_Schools = c(TRUE, TRUE, TRUE, FALSE, TRUE),
  Area_km2 = c(10.2, 15.3, 8.7, 12.5, 11.0),
  Category = factor(c("urban", "residential", "residential", "suburban", "mixed"))
)

colnames(zvolen_df) <- c("District_Name", "Population_Size", "Has_Schools", "Area_km2", "Category")

str(zvolen_df)
summary(zvolen_df)

zvolen_df$Density <- zvolen_df$Population_Size / zvolen_df$Area_km2
zvolen_df
```

## Task 2: Subsetting & Missing Data in Your Mini-Universe

### 2.1. Querying Vectors

Let's start with the vectors you created in **Task 1**.

2.1.1. From your **Sequence vector** (the one with all your numbers):

- Select and print the 3rd element.
- Select and print elements 2 through 5.
- Select and print all values that are greater than 20 (or another number that makes sense for your data).

```{r}
# Select and print the 3rd element
sequence_vector[3]

# Select and print elements 2 through 5
sequence_vector[2:5]

# Select and print all values greater than 20
sequence_vector[sequence_vector > 20]

```

2.1.2. From your **Logical vector**:

- Select and print only the TRUE values. (Hint: `your_vector[your_vector]`).
- Use `which()` to get the indices (positions) of all the `TRUE` values.

```{r}
# Select and print only the TRUE values
logical_vec[logical_vec]

# Get indices (positions) of all TRUE values
which(logical_vec)
```

### 2.2. Querying Your "Universe" List

Use your main list from **Task 1**.

- Access and print the **description** string using the `$` operator
- Access and print your **3x3 matrix** using double brackets `[[ ]]`.
- Access and print the **second element** from your **vector element** within the list. (This will be a "list-within-a-list" access).

```{r}
# Access and print the description string using $
zvolen_list$description

# Access and print your 3x3 matrix using double brackets [[ ]]
zvolen_list[[7]]   # or zvolen_list$matrix

# Access and print the 2nd element from your numeric vector inside the list
zvolen_list$numeric_vector[2]

```

### 2.3. Investigating Your Data Frame

Use the **data frame** you created in **Task 1**.

- Print the **first 3 rows**. (Use `head()`.)
- Print the value from the **4th row, 1st column**.
- Print only the column that contains your logical values (e.g., `is_habitable`) as a vector.

**Logical Subsetting**: Create a new data frame `df_subset` that contains only the rows that meet a specific logical condition.

```{r}
# Print the first 3 rows
head(zvolen_df, 3)

# Print the value from the 4th row, 1st column
zvolen_df[4, 1]

# Print only the logical column (Has_Schools)
zvolen_df$Has_Schools


# Logical Subsetting: keep only districts with population > 8000
zvolen_subset <- zvolen_df[zvolen_df$Population_Size > 8000, ]
zvolen_subset

# %in% Subsetting: select districts with "urban" or "residential" category
zvolen_subset_2 <- zvolen_df[zvolen_df$Category %in% c("urban", "residential"), ]
zvolen_subset_2

```

### 2.4. Handling "Corrupted" Data (`NA`)

Create a copy of your original data frame (e.g., `df_corrupted <- your_df`).

**Introduce 3 `NA`s** into your df_corrupted at specific locations.

- Example: `df_corrupted[2, 3] <- NA`

**Find `NAs`**:

- Use `sum(is.na(df_corrupted)`) to find the **total number** of missing values.
- Use `colSums(is.na(df_corrupted))` to see how many `NA`s are in **each column**.

**Calculate with `NA`s**:

- Calculate the `mean()` of the numeric column where you added an `NA`.

**Filter `NA`s**:

- Use `na.omit(df_corrupted)` to create a new, "clean" data frame that contains only the complete rows.

```{r}
# Create a copy of your original data frame
zvolen_corrupted <- zvolen_df

# Introduce 3 NA values in specific locations
zvolen_corrupted[2, 2] <- NA     # Population_Size for East
zvolen_corrupted[4, 3] <- NA     # Has_Schools for North
zvolen_corrupted[5, 5] <- NA     # Category for South

# Find total number of missing values
sum(is.na(zvolen_corrupted))

# Find how many NAs per column
colSums(is.na(zvolen_corrupted))

# Calculate mean of Population_Size (with NAs)
mean(zvolen_corrupted$Population_Size)

# Calculate mean ignoring NAs
mean(zvolen_corrupted$Population_Size, na.rm = TRUE)

# Remove rows with NAs
zvolen_clean <- na.omit(zvolen_corrupted)
zvolen_clean
```

### 2.5. Bonus Challenge: `which()`

Use `which.max()` or `which.min()` to find the **row number** (index) of the observation with the highest or lowest value in one of your numeric columns.

Use that index to print the row with that observation.

- Example: `highest_pop_index <- which.max(your_df$population)`
- Example: `your_df[highest_pop_index, ]`

```{r}
# Find the district with the highest population
highest_pop_index <- which.max(zvolen_df$Population_Size)
zvolen_df[highest_pop_index, ]

# Find the district with the lowest population
lowest_pop_index <- which.min(zvolen_df$Population_Size)
zvolen_df[lowest_pop_index, ]
```
## Task 3: Expanding Your Universe

Your "mini-universe" in Tasks 1 & 2 was a simulation. Now, it's time to connect it to the real world.

Your mission is to find a real dataset that expands on your theme, import it into R, and explore it using the dplyr package.

### 3.1. The Quest: Find Your Data

Find one simple, real-world dataset that fits your theme.

- **Where to look:** Kaggle, TidyTuesday, data.gov, or other public data portals.

- **Format:** Any format you are able to read. For simplicity it is proposed to use **textual (csv, txt)** or **Excel** format.

- **Requirements:**

    - At least 50 rows.

    - At least 4 columns.

    - A mix of numeric and character/factor data.

- **Action:**

    1. Download the data file and place it in your R project directory.

    2. In your task file add a line with the link to the data source (e.g., Data from: https://...).
    
### 3.2. The Import: Load Your Data

1. **Install** `tidyverse`: If you haven't, run `install.packages("tidyverse")` in your console (just once).

2. **Load libraries:** At the top of your script, add `library(dplyr)` and `library(readr)` (`library(readxl)` in case of Excel file format).

3. **Read the data:** Use `read_csv()` (or `read_excel()`) to load your dataset into a new data frame.

```{r}
library(tidyverse)

# Load the dataset
zoo <- read_csv("zoo.csv")

# Inspect structure
glimpse(zoo)
summary(zoo)
head(zoo, 10)
```

### 3.3. The Interrogation: Explore with `dplyr`

This is the core of the task. Your goal is to **ask and answer 5 interesting questions** about your new dataset.

You must write the question in plain English (as a comment), followed by the `dplyr` code that answers it.

#Question 1 — How many animals have more than 4 legs?

```{r}
animals_many_legs <- zoo %>%
  filter(legs > 4) %>%
  select(animal_name, legs)

animals_many_legs
```


#Question 2 — Create a new variable for “is_big_legged” (TRUE if legs ≥ 4)

```{r}
zoo2 <- zoo %>%
  mutate(is_big_legged = legs >= 4)

head(zoo2)
```

#Question 3 — What are the TOP 10 heaviest animals by number of legs?

```{r}
top10_legs <- zoo %>%
  arrange(desc(legs)) %>%
  head(10)

top10_legs
```

#Question 4 — How many animals belong to each class?

```{r}
class_counts <- zoo %>%
  group_by(class_type) %>%
  summarise(n = n())

class_counts

```

#Question 5 — Average number of legs per class

```{r}
legs_by_class <- zoo %>%
  group_by(class_type) %>%
  summarise(avg_legs = mean(legs))

legs_by_class
```

## Task 4. Building a "Toolkit" for Your Universe (Functions)

In Task 3, you performed a "one-off" analysis of your new dataset. In the real world, you'll want to run that analysis again and again.

The best way to do this is by capturing your code inside **functions**. A function is a reusable "tool" that you build to perform a specific job.

Your mission is to build a "toolkit" of functions that can analyse the dataset you found in Task 3.

**Important:** You must use the **same dataset** and theme you used in Task 3.

### 4.1. The "Greeting" Function (A Simple Start)

Let's start with the basics. Write a simple function that prints a welcome message for your "universe".

```{r}
welcome_user <- function(name) {
  paste("Welcome to Sofia's zoo data analysis universe,", name, "!")
}

welcome_user("Sofia")

```


### 4.2. The "Loading" Function (A Good Practice)

In Task 3, you wrote code to load and inspect your data. Let's turn that into a function. This is a very common and useful practice.

```{r}
load_and_inspect <- function(path) {
  data <- read_csv(path)
  cat("---- Structure ----\n")
  print(str(data))
  cat("\n---- Summary ----\n")
  print(summary(data))
  return(data)
}

zoo_loaded <- load_and_inspect("zoo.csv")

```

### 4.3. The "Query" Function (The Core of the Task)

This is the most important step. You will turn one of your `dplyr` queries from **Task 3** into a reusable function.

Look back at your Task 3 code. You probably wrote something like this:

  - `my_data |> filter(population > 1000000)`

  - `my_data |> filter(planet_type == "Gas Giant")`

That 1000000 or "Gas Giant" is a "magic number" or "magic string". A function lets you turn it into a parameter.

```{r}
query_animals <- function(data, column, value) {
  data %>%
    filter(.data[[column]] > value)
}

query_animals(zoo, "legs", 4)
```

### 4.4. The "Reporting" Function

Let's combine everything. Write a single function that calculates several key statistics for a specific **group** or **category** in your data.

```{r}
generate_report <- function(data) {
  cat("===== Zoo Dataset Report =====\n\n")
  
  cat("Number of rows:", nrow(data), "\n")
  cat("Number of columns:", ncol(data), "\n\n")
  
  # Animal with the most legs
  max_legs_row <- data[which.max(data$legs), ]
  cat("Animal with the MOST legs:\n")
  print(max_legs_row$animal_name)
  cat("Legs:", max_legs_row$legs, "\n\n")
  
  # Count per class
  cat("Animals per class:\n")
  print(data %>% group_by(class_type) %>% summarise(count = n()))
  
  cat("\n===== End of Report =====\n")
}

generate_report(zoo)
```

## Task 5. Visualising Your Universe (Base R & ggplot2)

You have found your data (Task 3) and built tools to analyse it (Task 4). Now, you must present your findings to customers.

Your goal is to create four specific visualizations based on the real-world dataset you imported in Task 3.

**Prerequisites:**

  - Use the dataset you found and cleaned in Task 3.

  - Make sure `library(ggplot2)` is loaded.

  - **Crucial:** Every plot must have a **Title**, **x-axis label**, and **y-axis label** that make sense for your specific theme.
  
```{r}
library(ggplot2)
library(dplyr)
zvolen_df <- data.frame(
  District_Name = c("Center", "East", "West", "North", "South"),
  Population_Size = c(12000, 9500, 7800, 5500, 7000),
  Has_Schools = c(TRUE, TRUE, TRUE, FALSE, TRUE),
  Area_km2 = c(10.2, 15.3, 8.7, 12.5, 11.0),
  Category = c("urban", "residential", "residential", "suburban", "mixed")
) %>%
  mutate(Density = Population_Size / Area_km2)
```
  
### 5.1. The Quick Sketch (Base R Histogram)

Sometimes you need a quick look at the distribution of your data without loading heavy packages.

  1. Choose one **numeric variable** from your dataset (e.g., `price`, `temperature`, `speed`).

  2. Use the `hist()` function from base R.

  3. Customise the `main` title, `xlab`, and the bar `col` (colour).
  
```{r}
hist(
  zvolen_df$Population_Size,
  main = "Population Distribution Across Zvolen Districts",
  xlab = "Population Size",
  col = "lightblue",
  border = "white"
)
```
  
### 5.2. The Relationship Map (ggplot2 Scatter Plot)

Now, let's look for patterns using the industry-standard package, **ggplot2**.

  1. Choose **two numeric variables** to see if they are correlated.

  2. Create a scatter plot using `geom_point()`.
  
  3. Add `geom_smooth()` to add a relationship between variables to the plot.

  4. Add `labs()` to rename your axes and title.
  
```{r}
ggplot(zvolen_df, aes(x = Area_km2, y = Population_Size)) +
  geom_point(size = 3, alpha = 0.7, colour = "darkblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Relationship Between District Area and Population Size",
    x = "District Area (km²)",
    y = "Population Size"
  ) +
  theme_minimal()

```
  
### 5.3. The Comparison (ggplot2 Boxplot)

3. The Comparison (ggplot2 Boxplot)

Boxplots are excellent for comparing groups. Let's see how different categories in your universe perform.

  1. Choose a **categorical (or logical) variable** for the X-axis (e.g., city, planet_type, species).

  2. Choose a **numeric variable** for the Y-axis (e.g., price, gravity, lifespan).

  3. Create a plot using `geom_boxplot()`.

  4. **Bonus:** Use `fill` to colour the boxes by the same category.
  
```{r}
ggplot(zvolen_df, aes(x = Category, y = Density, fill = Category)) +
  geom_boxplot() +
  labs(
    title = "Population Density by District Category",
    x = "District Category",
    y = "Population Density (people per km²)"
  ) +
  theme_light() +
  theme(legend.position = "none")
```
  
### 5.4. The "Deep Dive" (Faceting or Custom)

The most powerful feature of `ggplot2` is faceting—creating small multiples of the same plot for different groups.

  1. Create a plot (histogram, bar plot, or scatter plot).

  2. Use `facet_wrap(~ your_category_variable)` to split the plot into a separate panel for each group in your data.

  3. This is incredibly useful for comparing distinct groups (e.g., separate charts for each City or Year).

```{r}

ggplot(zvolen_df, aes(x = Density)) +
  geom_histogram(bins = 8, fill = "coral", color = "black") +
  facet_wrap(~ Category) +
  labs(
    title = "Distribution of Population Density by District Category",
    x = "Population Density",
    y = "Count"
  ) +
  theme_minimal()


```

### 5.5. Final Step: Exporting (Saving Your Work)

A plot is useless if you can't put it in a report.

  1. Choose your favourite plot from the exercises above.

  2. Run the code to display it in R.

  3. Immediately run the following command to save it to your project folder:
  
```{r}
ggsave("zvolen_faceted_histogram.png", width = 8, height = 6)
```
  
  