---
title: "Task1"
author: "Sofia"
date: "10/01/2024"
format:
  html:
    embed-resources: true
editor: source
---

# 1.1Atomic Types
Question description

Create the following variables related to your chosen object:
- 4 numeric values (e.g., population, temperature, size, rating)  
- 3 integer values (e.g., number of moons, buildings, inhabitants)  
- 2 logical values (e.g., has_water, is_inhabited)  
- 3 character values (e.g., name, region, category)

```{r}
# code for solving task
#integer 
population <- 39800
area <- 98  # square kilometers
districts <- 4

#charakter 
city_name <- "Zvolen"
city_old_name <- "Zolna"
region <- "Banska Bystrica Region" 

#numeric
average_temperature <- 10.1       # °C
city_budget_million <- 142.3      # million euros
annual_rainfall_mm <- 720.5       # mm per year
green_area_percent <- 36.8        # % of city covered by greenery

#logical 
has_hospital <- TRUE
is_capital <- FALSE
```

# 1.2Vectors
Create three different vectors:

1. Numeric vector combining all your numeric and integer variables.  
   *(Use c().)*

2. Sequence vector from any number between 10–50 to 100, Choose step between 2 and 10.  
   *(Use seq() or :.)*

3. Logical vector of length 10 with random TRUE/FALSE values using sample().

```{r}
#_code for solving task
# 1. Numeric vector
numeric_vector <- c(population, area, districts, average_temperature,
                    city_budget_million, annual_rainfall_mm, green_area_percent)
numeric_vector


# 2. Sequence vector
sequence_vector <- seq(5, 100, by = 20)
sequence_vector




# 3. Logical vector
logical_vec <- sample(c(TRUE, FALSE), size = 10, replace = TRUE)
logical_vec### 1.3 MATRIX

```


### 1.3 MATRIX
Create a 3×3 matrix with random numbers between -50 and 50  
(using sample() or runif()), and name its rows and columns according to your theme.

```{r}
#_code for solving task
set.seed(2024)
my_matrix <- matrix(sample(-50:50, 9, replace = TRUE), nrow = 3, ncol = 3)
rownames(my_matrix) <- c("Center", "East", "West")
colnames(my_matrix) <- c("Economy", "Education", "Tourism")
my_matrix

rowMeans(my_matrix)
colMeans(my_matrix)# 1.4 LIST
```
# 1.4 LIST
Create a list that combines all your previous objects:  
atomic variables, vectors, and matrix.
Add one extra element — a short description of your universe as a string.
```{r}
#_code for solving task
zvolen_list <- list(
  city_name = city_name,
  region = region,
  population = population,
  numeric_vector = numeric_vector,
  sequence_vector = sequence_vector,
  logical_vec = logical_vec,
  matrix = my_matrix,
  description = "This is the data universe of Zvolen a historic Slovak city full of green spaces, education, and culture."
)
zvolen_list
```

{r}
















# 1.5 Factor
Create a factor variable representing categories within your universe, e.g.:












- "low", "medium", "high"  
- "infant", "child", "adult"  
- "poor", "average", "rich"

Make sure you have at least 5 observations total.  
Then use table() to show how many objects fall into each category.
```{r}
#code_for solving task
development_stage <- factor(c("low", "medium", "high", "medium", "high", "low", "medium"))
table(development_stage)# 1.6 Data Frame
```

# 1.6 Data Frame
Create a data frame related to your object that includes:

1. 5–8 observations (rows)  
2. At least 3 columns of different data types  
   (e.g., name, numeric value, logical flag, category)
   dka
   dsds
   
   
   
   
   
   
   
   
   
   
```{r}
#_code for solving task
zvolen_df <- data.frame(
  District = c("Center", "East", "West", "North", "South"),
  Population = c(12000, 9500, 7800, 5500, 7000),
  Has_Schools = c(TRUE, TRUE, TRUE, FALSE, TRUE),
  Area_km2 = c(10.2, 15.3, 8.7, 12.5, 11.0),
  Category = factor(c("urban", "residential", "residential", "suburban", "mixed"))
)

colnames(zvolen_df) <- c("District_Name", "Population_Size", "Has_Schools", "Area_km2", "Category")

str(zvolen_df)
summary(zvolen_df)

zvolen_df$Density <- zvolen_df$Population_Size / zvolen_df$Area_km2
zvolen_df
```
